---
title: "Reproducible Research Project 1"
author: "Anis Rahman"
date: "03/20/2018"
output:
  md_document:
    variant: markdown_github
---


## Loading and preprocessing the data

Code is automated and set the current directory as working directory 
to download the required dataset

```{r,echo=TRUE}
library(dplyr)
library(ggplot2)
URL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"

if (!file.exists("F2activity.zip")){
    print("Download in progress")
    download.file(URL,destfile = "F2activity.zip")
    print("Download complete .....")
    print("File Unzip in progress .....")
    unzip("F2activity.zip")
    print("Unzip complete, file should reside in Working directory")
    print("reading file to dataframe")
    activityDf <- read.csv("activity.csv")
    print("Completed")
}else{
    print("reading file to dataframe")
    activityDf <- read.csv("activity.csv")
    print("Completed")
}

```

## What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day

```{r,echo=TRUE}
daily_steps <- activityDf %>% group_by(date)%>%summarise(total.steps =  sum(steps,na.rm = T))

head(daily_steps,10)

```

2. If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day
```{r,echo=TRUE}

act.Hist<-ggplot(daily_steps,aes(x=total.steps,fill="blue"))+geom_histogram(fill="red",binwidth=500)
act.Hist+labs(title="Daily Steps Histogram",x="Steps",y="Frequency")

```

## What is the average daily activity pattern?

```{r,echo=TRUE}


```

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
